### 2020-04-10 thesis/content/chapter-appendix.tex
continued with real-world section

@@

\paragraph{PROTEINS and D\&D}
The graphs in the PROTEINS dataset represent proteins~\cite{Borgwardt2005a}.
Each vertex corresponds to a so-called \ac{sse}, i.e.\ a certain molecular substructure.
An edge encodes either that two \acp{sse} are neighbors in the protein's amino-acid sequence or that those \acp{sse} are close to each other in 3D space.
Each protein graph is classified by whether it is an enzyme or not.
Very similarly, the D\&D dataset~\cite{Dobson2003} is also about classifying proteins as enzymes or non-enzymes but it uses a different selection of vertex features.

@@

>>>>

To tune the hyperparameters of the evaluated models, we used a regular grid search.
Depending on the type of model, different sets of hyperparameter configurations $\Theta$ were used.

\paragraph{Graph Kernels}
As described in \cref{sec:eval:setup}, we used the \ac{svm} classifier from \citetitle{SKL} to evaluate the graph kernel approaches.
We tuned only the regularization parameter \texttt{C} of this classifier;
the evaluated values are $\mathtt{C} \in \{ 1, \num{1e-1},\allowbreak \num{1e-2},\allowbreak \num{1e-3},\allowbreak \num{1e-4} \}$.
All other parameters were left at the default setting (using \texttt{scikit-learn 0.22.1}).

\paragraph{Baseline and \ac{gin}}
For the evaluation of the structure unaware baseline learner and \ac{gin}, we used the same hyperparameter configurations as \citet{Errica2020}.
We therefore refer to their work for a complete list of the tuned hyperparameters for those models.

\paragraph{2-\acs{gnn} and 2-\acs{wl}-\acs{gnn}}
We evaluated our implementations of 2-\acsp{gnn} as well as 2-\acs{wl}-\acsp{gnn} on the grid spanned by the following hyperparameter values:
\begin{itemize}[itemsep=2pt,parsep=2pt]
	\item \textbf{Number of convolutional layers $T \in \{ 3, 5 \}$:}
		This parameter describes only the depth of the stack of convolutional layers.
		The \ac{mlp} after the pooling layer is always configured with a single hidden layer.
	\item \textbf{Layer width $d \in \{ 32, 64 \}$:}
		This parameter describes the output dimensionalities $d = d^{(1)} = \cdots = d^{(T)}$ of the convolutional layers and (if applicable) also the hidden layer width of the final \ac{mlp} after the pooling layer.
	\item \textbf{Jumping knowledge $\mathrm{JK} \in \{ \mathtt{true}, \mathtt{false} \}$:}
		\citet{Xu2018a} have demonstrated that it can be advantageous for graph classification to pass the outputs of convolutional layers to their successors.
		We incorporate this idea by passing the input feature vectors not only to the first convolution layer but to all convolution layers through concatenation iff.\ $\mathrm{JK} = \mathtt{true}$.
	\item \textbf{Learning rate $\eta \in \{ \num{1e-2}, \num{1e-3}, \num{1e-4} \}$} of the Adam optimizer.
	\item \textbf{Activation functions $\sigma$ and $\sigma_{\Gamma}$} are set to the standard logistic function.
\end{itemize}

@@

\paragraph{PROTEINS and D\&D}
The graphs in both the PROTEINS dataset~\cite{Borgwardt2005a} as well as the D\&D dataset~\cite{Dobson2003} represent proteins.
Each vertex corresponds to a so-called \ac{sse}, i.e.\ a certain molecular substructure.
An edge encodes either that two \acp{sse} are neighbors in the protein's amino-acid sequence or that those \acp{sse} are close to each other in 3D space.
Each protein graph is classified by whether it is an enzyme or not.
The main difference between the two datasets is their selection of vertex features/labels.

@@

Due to the relatively small sizes of the evaluated benchmark datasets, the variance of the test accuracies across different folds is quite large.
When directly comparing the mean accuracies of two learners, it is therefore often impossible to tell whether one consistently outperforms the other.
We therefore now list the mean and standard deviations of the fold-wise test accuracy differences of all pairs of learners for all datasets.
This effectively removes the variance introduced by ``easy'' and ``hard'' folds on which all learners might tend to perform consistently better/worse.

In the following \crefrange{tbl:appendix:diff-triangle}{tbl:appendix:diff-imdb} (see \cpagerefrange{tbl:appendix:diff-triangle}{tbl:appendix:diff-imdb}), the differences are computed as $\textit{row accuracy} - \textit{column accuracy}$.
For each row $i$ and column $j$ we highlight the corresponding cell $(i,j)$ in \textcolor{t_red}{red} or \textcolor{t_darkgreen}{green} iff.\ the learner~$i$ performs consistently \textcolor{t_red}{worse} (or \textcolor{t_darkgreen}{better} respectively) than $j$ with a significance level of $2\sigma$.
To compute the deltas for 2-\acs{wl}-\acsp{gnn}, we use the same neighborhood radii as in \cref{tbl:eval:synthetic,tbl:eval:real}, i.e.\ $r = 2$ for the synthetic triangle detection dataset and $r = 8$, $5$, $2$, $1$ and $4$ for NCI1, PROTEINS, D\&D, REDDIT and IMDB respectively.

{\captionsetup[table]{name=Matrix}%
\begin{table}[ht]
	\caption{Fold-wise accuracy delta means and standard deviations on the triangle dataset.}\label[mat]{tbl:appendix:diff-triangle}
	\centering
	\input{data/diffs/triangle.tex}
\end{table}
\begin{table}[ht]
	\caption{Fold-wise accuracy delta means and standard deviations on NCI1.}\label[mat]{tbl:appendix:diff-nci}
	\centering
	\input{data/diffs/nci.tex}
\end{table}
\begin{table}[ht]
	\caption{Fold-wise accuracy delta means and standard deviations on PROTEINS.}\label[mat]{tbl:appendix:diff-proteins}
	\centering
	\input{data/diffs/proteins.tex}
\end{table}
\begin{table}[ht]
	\caption{Fold-wise accuracy delta means and standard deviations on D\&D.}\label[mat]{tbl:appendix:diff-dd}
	\centering
	\input{data/diffs/dd.tex}
\end{table}
\begin{table}[ht]
	\caption{Fold-wise accuracy delta means and standard deviations on REDDIT.}\label[mat]{tbl:appendix:diff-reddit}
	\centering
	\input{data/diffs/reddit.tex}
\end{table}
\begin{table}[ht]
	\caption{Fold-wise accuracy delta means and standard deviations on IMDB.}\label[mat]{tbl:appendix:diff-imdb}
	\centering
	\input{data/diffs/imdb.tex}
\end{table}%
}

### 2020-04-10 thesis/content/chapter-evaluation.tex
continued with real-world section

If we compare the real-world results with those for the synthetic triangle detection dataset, we do no longer observe the clear advantage of 2-\acs{wl}-\acsp{gnn} over the other approaches.
This indicates that the theoretical advantages of 2-\acs{wl} over 1-\acs{wl} are not necessarily relevant for the five evaluated problems.
Nonetheless, the test performance of 2-\acs{wl}-\acsp{gnn} is generally comparable to that of the other state-of-the-art learners with the exception of NCI1.
If we look at the enzyme detection problem (PROTEINS and D\&D), we observe that all evaluated approaches appear to be unable to leverage structural information for a significant improvement over the baseline learner.
In the social network datasets (REDDIT and IMDB) on the other hand, the structure aware methods clearly outperform the baseline.
This confirms the very similar results of \citet{Errica2020}.

>>>>

If we compare the real-world results with those for the synthetic triangle detection dataset, we do no longer observe the clear advantage of 2-\acs{wl}-\acsp{gnn} over the other approaches.
This indicates that the theoretical advantages of 2-\acs{wl} over 1-\acs{wl} are not necessarily relevant for the five evaluated problems.
Nonetheless, the test performance of 2-\acs{wl}-\acsp{gnn} is generally comparable to that of the other state-of-the-art learners with the exception of NCI1,
where by comparable we mean that the performance of the evaluated 2-\acs{wl}-\acs{gnn} models is within the $2\sigma$ confidence interval of the best evaluated model even when compared fold-by-fold (see \cref{sec:appendix:fold-diffs}).

If we look at the enzyme detection problem (PROTEINS and D\&D), we observe that all evaluated approaches appear to be unable to leverage structural information for a significant improvement over the baseline learner (see \cref{tbl:appendix:diff-proteins,{tbl:appendix:diff-dd}}).
In the social network datasets (REDDIT and IMDB) on the other hand, the structure aware methods clearly outperform the baseline (see \cref{tbl:appendix:diff-imdb}).
This confirms the very similar results of \citet{Errica2020}.

\begin{figure}[ht]
	\centering
	\begin{tikzpicture}
		\begin{axis}[
			width=0.33\linewidth,
			height=0.3\linewidth,
			try min ticks=6,
			xmin=1, xmax=5,
			ymin=70, ymax=100,
			label style={font=\tiny},
			tick label style={font=\tiny},
			ymajorgrids,
			ytick style={draw=none},
			axis line style={gray},
			mark size={1.5pt}
		]
			\addplot [color=t_blue, only marks, mark=*]
			plot[error bars/.cd, y dir=both, y explicit]
			table [x=T, y=nciTrainMean, y error=nciTrainStd, col sep=semicolon] {data/wlst_depths.csv};
			\addplot [color=t_red, only marks, mark=square*]
			plot[error bars/.cd, y dir=both, y explicit]
			table [x=T, y=nciTestMean, y error=nciTestStd, col sep=semicolon] {data/wlst_depths.csv};
		\end{axis}
	\end{tikzpicture}
	\caption{test}\label{fig:eval:wlst-depths}
\end{figure}

### 2020-04-10 thesis/content/chapter-ltd.tex
continued with real-world section

	W.l.o.g.\ we define the initial colors of the loop vertices as $\chi^{(0)}(v) = \colorlabel{t_blue}{A}$ for all $v \in L_G \cup L_H$. % chktex 25
	The initial colors of the edge vertices are defined as $\chi^{(0)}(e) = \textcolor{t_red}{\texttt{B}}$ for all $e \in E_G \cup E_H$. % chktex 25
	Note that each loop vertex $\ldblbrace v_i, v_i \rdblbrace$ with $v_i \in \mathcal{V}_G \cup \mathcal{V}_H$ has $d$ neighbors, the edges incident to $v_i$.
	Similarly each edge vertex $\ldblbrace v_i, v_j \rdblbrace$ has $2d$ neighbors, two of which are the loop vertices $\ldblbrace v_i, v_i \rdblbrace$ and $\ldblbrace v_j, v_j \rdblbrace$ with the remaining $2d - 2$ neighbors corresponding to the edges that are incident to $e_{ij}$.
	This is illustrated in \cref{fig:ltd:edge-neighborhood}.

	After one color refinement step we get $\chi^{(1)}(v) = h(\colorlabel{t_blue}{A}, \ldblbrace \underbrace{\textcolor{t_red}{\texttt{B}}, \dots, \textcolor{t_red}{\texttt{B}}}_{d\text{ times}} \rdblbrace) \eqqcolon \textcolor{t_blue}{\texttt{C}}$ for all loop vertices $v \in  L_G \cup L_H$ and $\chi^{(1)}(e) = h(\textcolor{t_red}{\texttt{B}}, \ldblbrace \colorlabel{t_blue}{A}, \colorlabel{t_blue}{A}, \underbrace{\textcolor{t_red}{\texttt{B}}, \dots, \textcolor{t_red}{\texttt{B}}}_{\mathclap{2d - 2 \text{ times}}} \rdblbrace) \eqqcolon \textcolor{t_red}{\texttt{D}}$. % chktex 25
	This means that $\chi^{(0)}$ and $\chi^{(1)}$ are identical up to the color substitutions $\colorlabel{t_blue}{A} \to \textcolor{t_blue}{\texttt{C}}$ and $\textcolor{t_red}{\texttt{B}} \to \textcolor{t_red}{\texttt{D}}$, i.e.\ $\chi^{(0)} \equiv \chi^{(1)}$, which in turn implies that 1-\ac{wl} terminates after one iteration. % chktex 25
	\Cref{lem:ltd:wl1-regular-edge-neighbor-limit} then directly follows since both $G^{\mathcal{E}}$ and $H^{\mathcal{E}}$ have $n$ vertices with the final color $\textcolor{t_blue}{\texttt{C}}$ and $\frac{nd}{2}$ vertices with the final color $\textcolor{t_red}{\texttt{D}}$, i.e.\ $G^{\mathcal{E}} \mathrel{\simeq_1} H^{\mathcal{E}}$. % chktex 25
\end{proof}
\begin{figure}[ht]
	\centering
	\includegraphics[width=\linewidth]{gfx/ltd/edge-neighborhood.pdf}
	\caption{
		Illustration of the edge neighborhood graphs of two 2-regular graphs of size 6.
	}\label{fig:ltd:edge-neighborhood}
\end{figure}
\begin{prop}\label{prop:ltd:2gnn-regular-limit}
	A 2-\acs{gnn} cannot distinguish regular graphs of the same size and therefore has a lower discriminative power than 2-\ac{wl}.
\end{prop}
\begin{proof}
	The proposition directly follows from \cref{prop:ltd:2gnn-wl1-limit}, \cref{lem:ltd:wl1-regular-edge-neighbor-limit} and the fact that 2-\acs{wl} is able to distinguish most regular graphs~\cite[cor.~1.8.6]{Immerman1990}.
\end{proof}

@@

	Let $\varphi: \mathbb{R}^{d^{(T, 2 + K)}} \to \mathbb{R}^{d^{(T)}} \cup \{ \texttt{nil} \}$ be a function which maps the final 2-\acs{wl} feature vectors produced by $\mathit{Conv}_2$ to the output space of $\mathit{Conv}_1$ or the constant \texttt{nil}.
	Let $\mathit{Pool}_2\left(Z_G^{(T,2+K)}\right) \coloneqq \mathit{Pool}_1(\ldblbrace z_{ij}\, |\, z_{ij} = \varphi\left(Z_G^{(T,2+K)}[e_{ij}]\right) \land e_{ij} \in \mathcal{E}_{G^1} \land z_{ij} \neq \texttt{nil} \rdblbrace)$.
	\Cref{thm:ltd:wl2-simulation} then follows if there is a function $\varphi$ s.t.\ $\forall v_i \in \mathcal{V}_G: \mathit{Conv}_1(G)[v_i] = \varphi(\mathit{Conv}_2(G)[e_{ii}])$ and $\forall e_{ij} \in \mathcal{E}_G: \varphi(\mathit{Conv}_2(G)[e_{ij}]) = \texttt{nil}$.

@@

	Using \cref{inv:ltd:wl2-simulation:indicator,inv:ltd:wl2-simulation:feature} for $t = T$ we can therefore set
	\begin{align*}
		\varphi\left(Z_G^{(T,2+K)}[e_{ij}]\right) \coloneqq \begin{cases}
			{Z_G^{(T,2+K)}[e_{ij}]}_{2, \dots, (d^{T} + 1)} & \text{if ${Z_G^{(T,2+K)}[e_{ij}]}_1 = 1$} \\
			\texttt{nil} & \text{else}
		\end{cases}
		\text{.}
	\end{align*}
	By our previous definition of $\mathit{Pool}_2$, this in turn implies that $\mathit{Pool}_2(Z_G^{(T,2+K)}) = \mathit{Pool}_1(Z_G^{(T)}) \iff h_2(G) = h_1(G)$ which concludes the proof.
\end{proof}
\begin{cor}\label{cor:ltd:wl2-gnn-wl1-power}
	2-\acs{wl}-\acsp{gnn} have at least the same discriminative power as 1-\acs{wl}.
\end{cor}
\begin{proof}
	The corollary directly follows from the fact that 2-\acs{wl}-\acsp{gnn} can simulate \acp{gin} by \cref{thm:ltd:wl2-simulation} and the fact that \acp{gin} have the same discriminative power as 1-\acs{wl} by \cref{prop:related:gcnn-wl1-limit} (see \cpageref{prop:related:gcnn-wl1-limit}) because they use injective vertex neighborhood hashing functions~\cite{Xu2018}.
\end{proof}

>>>>

	W.l.o.g.\ we define the initial colors of the loop vertices as $\chi^{(0)}(v) = \colorlabel{t_blue}{A}$ for all $v \in L_G \cup L_H$. % chktex 25
	The initial colors of the edge vertices are defined as $\chi^{(0)}(e) = \colorlabel{t_red}{B}$ for all $e \in E_G \cup E_H$. % chktex 25
	Note that each loop vertex $\ldblbrace v_i, v_i \rdblbrace$ with $v_i \in \mathcal{V}_G \cup \mathcal{V}_H$ has $d$ neighbors, the edges incident to $v_i$.
	Similarly each edge vertex $\ldblbrace v_i, v_j \rdblbrace$ has $2d$ neighbors, two of which are the loop vertices $\ldblbrace v_i, v_i \rdblbrace$ and $\ldblbrace v_j, v_j \rdblbrace$ with the remaining $2d - 2$ neighbors corresponding to the edges that are incident to $e_{ij}$.
	This is illustrated in \cref{fig:ltd:edge-neighborhood}.

	After one color refinement step we get $\chi^{(1)}(v) = h(\colorlabel{t_blue}{A}, \ldblbrace \underbrace{\colorlabel{t_red}{B}, \dots, \colorlabel{t_red}{B}}_{d\text{ times}} \rdblbrace) \eqqcolon \colorlabel{t_blue}{C}$ for all loop vertices $v \in  L_G \cup L_H$ and $\chi^{(1)}(e) = h(\colorlabel{t_red}{B}, \ldblbrace \colorlabel{t_blue}{A}, \colorlabel{t_blue}{A}, \underbrace{\colorlabel{t_red}{B}, \dots, \colorlabel{t_red}{B}}_{\mathclap{2d - 2 \text{ times}}} \rdblbrace) \eqqcolon \colorlabel{t_red}{D}$. % chktex 25
	This means that $\chi^{(0)}$ and $\chi^{(1)}$ are identical up to the color substitutions $\colorlabel{t_blue}{A} \to \colorlabel{t_blue}{C}$ and $\colorlabel{t_red}{B} \to \colorlabel{t_red}{D}$, i.e.\ $\chi^{(0)} \equiv \chi^{(1)}$, which in turn implies that 1-\ac{wl} terminates after one iteration. % chktex 25
	\Cref{lem:ltd:wl1-regular-edge-neighbor-limit} then directly follows since both $G^{\mathcal{E}}$ and $H^{\mathcal{E}}$ have $n$ vertices with the final color $\colorlabel{t_blue}{C}$ and $\frac{nd}{2}$ vertices with the final color $\colorlabel{t_red}{D}$, i.e.\ $G^{\mathcal{E}} \mathrel{\simeq_1} H^{\mathcal{E}}$. % chktex 25
\end{proof}
\begin{figure}[ht]
	\centering
	\includegraphics[width=\linewidth]{gfx/ltd/edge-neighborhood.pdf}
	\caption{
		Illustration of the edge neighborhood graphs of two 2-regular graphs of size 6.
	}\label{fig:ltd:edge-neighborhood}
\end{figure}
\begin{prop}\label{prop:ltd:2gnn-regular-limit}
	A 2-\acs{gnn} cannot distinguish regular graphs of the same size and therefore has a lower discriminative power than 2-\ac{wl}.
\end{prop}
\begin{proof}
	The proposition directly follows from \cref{prop:ltd:2gnn-wl1-limit}, \cref{lem:ltd:wl1-regular-edge-neighbor-limit} and the fact that 2-\acs{wl} is able to distinguish most regular graphs~\cite[cor.~1.8.6]{Immerman1990}.
\end{proof}

@@

	Let $\varphi: \mathbb{R}^{d^{(T, 2 + K)}} \to \mathbb{R}^{d^{(T)}} \cup \{ \mathtt{nil} \}$ be a function which maps the final 2-\acs{wl} feature vectors produced by $\mathit{Conv}_2$ to the output space of $\mathit{Conv}_1$ or the constant \texttt{nil}.
	Let $\mathit{Pool}_2\left(Z_G^{(T,2+K)}\right) \coloneqq \mathit{Pool}_1(\ldblbrace z_{ij}\, |\, z_{ij} = \varphi\left(Z_G^{(T,2+K)}[e_{ij}]\right) \land e_{ij} \in \mathcal{E}_{G^1} \land z_{ij} \neq \mathtt{nil} \rdblbrace)$.
	\Cref{thm:ltd:wl2-simulation} then follows if there is a function $\varphi$ s.t.\ $\forall v_i \in \mathcal{V}_G: \mathit{Conv}_1(G)[v_i] = \varphi(\mathit{Conv}_2(G)[e_{ii}])$ and $\forall e_{ij} \in \mathcal{E}_G: \varphi(\mathit{Conv}_2(G)[e_{ij}]) = \mathtt{nil}$.

@@

	Using \cref{inv:ltd:wl2-simulation:indicator,inv:ltd:wl2-simulation:feature} for $t = T$ we can therefore set
	\begin{align*}
		\varphi\left(Z_G^{(T,2+K)}[e_{ij}]\right) \coloneqq \begin{cases}
			{Z_G^{(T,2+K)}[e_{ij}]}_{2, \dots, (d^{T} + 1)} & \text{if ${Z_G^{(T,2+K)}[e_{ij}]}_1 = 1$} \\
			\mathtt{nil} & \text{else}
		\end{cases}
		\text{.}
	\end{align*}
	By our previous definition of $\mathit{Pool}_2$, this in turn implies that $\mathit{Pool}_2(Z_G^{(T,2+K)}) = \mathit{Pool}_1(Z_G^{(T)}) \iff h_2(G) = h_1(G)$ which concludes the proof.
\end{proof}
\begin{cor}\label{cor:ltd:wl2-gnn-wl1-power}
	2-\acs{wl}-\acsp{gnn} have at least the same discriminative power as 1-\acs{wl}.
\end{cor}
\begin{proof}
	The corollary directly follows from the fact that 2-\acs{wl}-\acsp{gnn} can simulate \acp{gin} by \cref{thm:ltd:wl2-simulation} and the fact that \acp{gin} have the same discriminative power as 1-\acs{wl} by \cref{prop:related:gcnn-wl1-limit} (see \cpageref{prop:related:gcnn-wl1-limit}) because they use injective vertex neighborhood hashing functions~\cite{Xu2018}.
\end{proof}
